Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-19 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Text edits to Methods: Order parameters from simulations

\author{Alexandru Botan}
%\affiliation{The authors are listed in alphaphetical order.}
%\affiliation{The author list is not completed.}
\affiliation[Lyon CNRS]{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
%\author{Andrea Catte}
%\affiliation{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela-Rosales}
\affiliation[Mexico]{Departamento de F\'isica, Centro de Investigaci\'on y de Estudios Avanzados del IPN, Apartado Postal 14-740, 07000 M\'exico D.F., M\'exico}
\author{Patrick F. J. Fuchs}
\affiliation[CNRS Paris]{Institut Jacques Monod, CNRS, Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
%\alsoaffiliation[Diderot University]
%{Institut Jacques Monod, CNRS UMR 7592, Universit{\'e} Paris Diderot, Sorbonne Paris Cit{\'e}, Paris, France}
\author{Matti Javanainen}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Matej Kandu\v{c}}
\affiliation[Freie Universit\"{a}t Berlin] 
{Fachbereich Physik, Freie Universitat Berlin, Berlin, Germany}
\author{Waldemar Kulig}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\affiliation[Kyoto University]
{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
\author{Claire Loison}
\affiliation[Lyon CNRS]{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
\author{Alexander Lyubartsev}
\affiliation[Stockholm University]
{Division of Physical Chemistry, Department of Materials and Environmental Chemistry, Stockholm University, S-106 91 Stockholm, SWEDEN}
\author{Markus S. Miettinen}
\affiliation[Freie Universit\"{a}t Berlin] 
{Fachbereich Physik, Freie Universitat Berlin, Berlin, Germany}
\author{Luca Monticelli}
\affiliation[IBCP] 
{Institut de Biologie et Chimie des Prot\'eines (IBCP), CNRS UMR 5086, Lyon, France}
\author{Jukka M{\"a}{\"a}tt{\"a}}
\affiliation[Aalto University]
{Aalto University, Espoo, Finland}
\author{O. H. Samuli Ollila} 
\email{samuli.ollila@aalto.fi.}
\affiliation[Aalto University]
{Aalto University, Espoo, Finland}
\author{Marius Retegan}
\affiliation[Max Planck]
{Max Planck Institute for Chemical Energy Conversion, Mulheim an der Ruhr, Germany}
\author{Tomasz Rog}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Hubert Santuz}
\affiliation[INSERM]
{INSERM, UMR\_S 1134, DSIMB, Paris, France}
\alsoaffiliation[Diderot]
{Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, UMR\_S 1134, Paris, France}
\alsoaffiliation[INTS]
{Institut National de la Transfusion Sanguine (INTS), Paris, France}
\alsoaffiliation[Labex]
{Laboratoire d'Excellence GR-Ex, Paris, France}
\author{Joona Tynkkynen}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}

@@

\subsection{Order parameters from simulations}
The order parameters from simulations were calculated directly using the definition as in Eq.~\ref{orderP}.
For the united atom models the hydrogen positions were generated 
in the trajectories post-simulationally using the positions of the heavy atoms and the known hydrocarbon geometries.
For the statistical error estimates, the time average of order parameters were first calculated separately
for each lipid in the system. Then it was assumed that different lipids are statistically independent 
entities (which should be the case in fluid phase) and the error of the mean for the average over individual 
lipids in the system was calculated and used as error bar for the order parameters.

It has been recently pointed out that the sampling of individual dihedral angles might be very
slow compared to the typical simulation timescales~\cite{vogel12}. On the other hand, another recent
study shows that the slowest rotational correlation functions of a C--H bond (g$_1$) reaches a plateau (S$_{CH}^2$)
after 200~ns in the Berger-POPC-07~\cite{ollila07a} model, and that the dynamics of this segment is significantly too slow in simulations
compared to the experiments~\cite{ferreira15}. In practise, less than 200~ns of simulation data is enough for the order parameter
calculation due to the average over different lipid molecules. In conclusion, if the sampling with typical simulation times
is not enough for the convergence of the order parameters, then the simulation models has significantly too slow dynamics.

>>>>

\author{Alexandru Botan}
%\affiliation{The authors are listed in alphaphetical order.}
%\affiliation{The author list is not completed.}
\affiliation[Lyon CNRS]{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
%
%\author{Andrea Catte}
%\affiliation{University of East Anglia, Norwich, United Kingdom}
\author{Fernando Favela-Rosales}
\affiliation[Mexico]{Departamento de F\'isica, Centro de Investigaci\'on y de Estudios Avanzados del IPN, Apartado Postal 14-740, 07000 M\'exico D.F., M\'exico}
\author{Patrick F. J. Fuchs}
\affiliation[CNRS Paris]{Institut Jacques Monod, CNRS, Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, Paris, France}
%\alsoaffiliation[Diderot University]
%{Institut Jacques Monod, CNRS UMR 7592, Universit{\'e} Paris Diderot, Sorbonne Paris Cit{\'e}, Paris, France}
\author{Matti Javanainen}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Matej Kandu\v{c}}
\affiliation[Freie Universit\"{a}t Berlin] 
{Fachbereich Physik, Freie Universitat Berlin, Berlin, Germany}
\author{Waldemar Kulig}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Antti Lamberg}
\affiliation[Kyoto University]
{Department of Chemical Engineering, Kyoto University, Kyoto, Japan}
\author{Claire Loison}
\affiliation[Lyon CNRS]{Institut Lumi\`ere Mati\`ere, UMR5306 Universit\'e Lyon 1-CNRS, Universit\'e de Lyon 69622 Villeurbanne, France}
\author{Alexander Lyubartsev}
\affiliation[Stockholm University]
{Division of Physical Chemistry, Department of Materials and Environmental Chemistry, Stockholm University, S-106 91 Stockholm, SWEDEN}
\author{Markus S. Miettinen}
\affiliation[Freie Universit\"{a}t Berlin] 
{Fachbereich Physik, Freie Universit\"at Berlin, Berlin, Germany}
\author{Luca Monticelli}
\affiliation[IBCP] 
{Institut de Biologie et Chimie des Prot\'eines (IBCP), CNRS UMR 5086, Lyon, France}
\author{Jukka M{\"a}{\"a}tt{\"a}}
\affiliation[Aalto University]
{Aalto University, Espoo, Finland}
\author{O. H. Samuli Ollila} 
\email{samuli.ollila@aalto.fi.}
\affiliation[Aalto University]
{Aalto University, Espoo, Finland}
\author{Marius Retegan}
\affiliation[Max Planck]
{Max Planck Institute for Chemical Energy Conversion, Mulheim an der Ruhr, Germany}
\author{Tomasz Rog}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}
\author{Hubert Santuz}
\affiliation[INSERM]
{INSERM, UMR\_S 1134, DSIMB, Paris, France}
\alsoaffiliation[Diderot]
{Universit\'e Paris Diderot, Sorbonne Paris Cit\'e, UMR\_S 1134, Paris, France}
\alsoaffiliation[INTS]
{Institut National de la Transfusion Sanguine (INTS), Paris, France}
\alsoaffiliation[Labex]
{Laboratoire d'Excellence GR-Ex, Paris, France}
\author{Joona Tynkkynen}
\affiliation[Tampere University of Technology]
{Department of Physics, Tampere University of Technology, Tampere, Finland}

@@

\subsection{Order parameters from simulations}
The order parameters from simulations were calculated directly using the definition of Eq.~\ref{orderP}.
For the united atom models, the hydrogen positions were generated 
post-simulationally from the positions of the heavy atoms and the known hydrocarbon geometries.
The statistical error was estimated based on the assumption that different lipids are statistically independent 
entities (which should be the case in fluid phase):
The time average of a given order parameter was first calculated separately for each lipid,
and the standard error of the mean over these time averages then taken as the error bar for this order parameter.

It has been pointed out that the sampling of individual dihedral angles might be very
slow compared to the typical (100~ns) simulation timescales~\cite{vogel12}.
After 200~ns, however, even the slowest rotational correlation function of a C--H bond  (g$_1$) reaches
a plateau ($S_\mathrm{CH}^2$) in the Berger-POPC-07 model~\cite{ollila07a}---and, notably,
the dynamics of this segment have been shown to be significantly slower in simulations
than in experiments~\cite{ferreira15}.
In practise,
due to averaging over different lipid molecules,
less than 200~ns of simulation data should be enough for the order parameter calculation;
if the sampling within typical simulation times
is not enough for the convergence of the order parameters,
then the simulation model in question has unphysically slow dynamics.
